Imputation of Incomplete Multilevel Data with \pkg{mice}

### 2022-04-06 Manuscript/Imputation_of_Incomplete_Multilevel_Data.Rmd
Clean up manuscript file

---
documentclass: jss
author:
  - name: Hanne Oberman
    affiliation: Utrecht University
    address: |
      | Padualaan 14
      | 3584 CH Utrecht
    email: \email{h.i.oberman@uu.nl}
    url: https://hanneoberman.github.io/
  - name: "Johanna Munoz Avila"
    affiliation: University Medical Center Utrecht \AND
  - name: Valentijn de Jong
    affiliation: University Medical Center Utrecht 
  - name: Gerko Vink
    affiliation: Utrecht University \AND
  - name: Thomas Debray
    affiliation: University Medical Center Utrecht 
title:
  formatted: "Imputation of Incomplete Multilevel Data with \\pkg{mice}"
  plain:     "Imputation of Incomplete Multilevel Data with mice"
  short:     "Multilevel \\pkg{mice}"
abstract: >
  This is a tutorial paper on imputing incomplete multilevel data with \pkg{mice}. Footnotes in the current version show work in progress/under construction. The last section is not part of the manuscript, but purely for reminders. We aim to submit at JSS, so there is no word count limit ("There is no page limit, nor a limit on the number of figures or tables").
keywords:
  # at least one keyword must be supplied
  formatted: [missing data, multilevel, clustering, "\\pkg{mice}", "\\proglang{R}"]
  plain:     [missing data, multilevel, clustering, mice, R]
preamble: >
  \usepackage{amsmath}
header-includes: 
 - \usepackage{graphicx} 
 - \usepackage{mathtools}
 - \usepackage{ulem}
output: 
    rticles::jss_article:
      keep_tex: yes
      number_sections: yes
    # word_document: default
bibliography: ../References/multilevelmice.bib
editor_options: 
  chunk_output_type: inline
---

@@

# seed
set.seed(123)

# packages
library(tidyverse)
library(mice)
library(miceadds)
library(lme4)
library(broom.mixed)
# library(metamisc)
# library(GJRM)
# library(pan)

# functions
`%nin%` <- Negate(`%in%`)
round3 <- function(x){format(round(x, 3), nsmall = 3)}
icc <- function(formula, data){multilevel::ICC1(aov(formula, data))}
# icc for imputed data as function in mice
miceadds::source.all(path = "../R")

# plot parameters
plot_col <- mice:::mdc(1:2) %>% setNames(c("observed", "missing"))

# data
popcomp <- foreign::read.spss("../Data/popular.sav", to.data.frame = TRUE, use.value.labels = FALSE) %>% 
  mutate(school = SCHOOL,
         popular = POPULAR,
         sex = SEX,
         texp = TEXP,
         .keep = "none")
popmis <- mice::popmis[, c("school", "popular", "sex", "texp")] 
```

<!-- Missing data pattern with clustering -->
<!-- TODO: simplify argument in md pattern plot and/or network plot for md patterns (summarized per variable?) -->
<!-- TODO: test if there's a random distribution of missingness in the md pattern? -->
<!-- TODO: check spss missing data pie chart? -->
<!-- TODO: similarity of patterns over clusters -->
<!-- TODO: gradient (alpha) to show in how many of the clusters the row in the pattern occurs -->
<!-- TODO: 3 columns could be 3 boxplots, over these clusters we have a distribution of these statistics across clusters.  -->
<!-- TODO: look at vim -->

<!-- Comments Thomas: -->
<!-- .	We need to remind the reader about general knowledge about missing data: it usually leads to loss of power and bias; we distinguish between MCAR, MAR and MNAR in theory but in practice this distinction is less clear; multiple imputation is generally recommended to avoid bias (if data are MCAR or MAR, but also works reasonable in some situations where data are partially MNAR) -->
<!-- .	We need to clarify what is clustered data, and provide common types (IPDMA, registries with data from multiple hospitals/regions, and multicenter studies). You can directly refer to the examples you will use in the paper here, and build on the text you already had. -->
<!-- .	We need to clarify why clustering is relevant during imputation, and why this exposes the need for specialized imputation methods and more attention during their implemention ("though shall not simply run mice() on any incomplete dataset") -->
<!-- .	Maybe we should also emphasise that in more modern types of big data sources (E.g. EHR data ), measurement of data is often selective and MAR assumptions are less tenable.  -->

@@

In many contemporary data analysis efforts, some form of hierarchical or clustered data structures are recorded. In the simplest case, such a structure entails the nesting of units within clusters (e.g., students within school classes). More complex clustered structures may occur when there are multiple hierarchical levels (e.g., patients within hospitals within regions or countries), or when the clustering is non-nested (e.g., data from diverse settings and populations within large databases). The clustered structure of multilevel data should be taken into account when developing analysis models: 1) for the simple reason that groups of observations share some common variance, and 2) because ignoring multilevel structures can be harmful to the statistical inferences and introduce bias in estimators [@hox17]. There are many names for models that take clustering into account. Some popular examples are 'multilevel models', 'hierarchical models', 'mixed effect models' and 'random effect models'. Table \ref{tab:def} provides an overview of some key concepts in multilevel modeling.

<!-- Contemporary decision-making is  increasingly often data-driven and sometimes even entirely dictated by data. Although data relevant for decision makers were initially collected using small and well-designed studies, there has been a growing need to address more complex questions at a wider scale. This not only requires to collect larger amounts of data, but also from more diverse settings and populations. For example, [TD: introduce IPDMA here; eg. HIV studies Johanna is referring to]. Another example is the use of large databases with information from thousands or even millions of individuals from multiple cities, regions or even countries. Finally, a third situation arises when data are collected from multicenter studies [TD: maybe refer to the schools/classes example here]. For example, students may be clustered in classes in psychometrics research, or patients may be clustered in studies in individual patient data meta-analyses in medical research. A common characteristic in aforementioned examples is the presence of clustering. [TD: briefly explain what is clustering?] -->

<!-- Ignoring the clustered structure of such multilevel data can be harmful to the statistical inferences and introduce bias in estimators [@hox17].  -->
<!-- ~~Imagine a case where cross-level interactions between unit-level variables and cluster-level variables are present. The cluster to which a unit belongs may then influence the unit-level observations--and vice versa for each of the units that make up the cluster.~~  -->
<!-- These relations can and should be taken into account when developing analysis models for multilevel data for the simple reason that groups of observations share some common variance. -->
<!-- ~~The variability due to clustering is often measured by means of the intraclass coefficient (ICC). The ICC can be seen as the percentage of variance that can be attributed to the cluster-level, where a high ICC would indicate that a lot of variability is due to the cluster structure.~~  -->
<!-- Multilevel models typically accommodate for variability by including a separate group mean for each cluster. In addition to random intercepts, multilevel models can also include random effects and heterogeneous residual error variances across clusters [see e.g. @gelm06, @hox17 and @jong21].  -->

<!-- There are many names for models that take clustering into account. Some popular examples are 'multilevel models', 'hierarchical models', 'mixed effect models' and 'random effect models'. [TD: emphasize when/why we need to account for clustering in the analysis of clustered data. Why is the presence of clustering relevant when considering multiple imputation of missing data? e.g. distinction between systematically and sporadically missing data. But also: mechanisms of missing data (e.g. MCAR, MAR, MNAR) may differ between clusters. But also: relation between observed data may differ between clusters? When/ why should we avoid using traditional imputation methods? e.g. congeniality issues.] -->

\begin{table}[tb]
\caption{Concepts in multilevel modeling}
\label{tab:def}
\centering
\begin{tabular}{ll}
\hline
\textbf{Concept} & \textbf{Details}   \\
\hline
ICC                 & The variability due to clustering is often measured by means of the \\
                    & intraclass coefficient (ICC). The ICC can be seen as the percentage \\
                    & of variance that can be attributed to the cluster-level, where a high \\
                    & ICC would indicate that a lot of variability is due to the cluster \\
                    & structure. \\
Random intercept    & Multilevel models typically accommodate for variability by including \\
                    & a separate group mean for each cluster. In addition to random \\
                    & intercepts, multilevel models can also include random effects and \\
                    & heterogeneous residual error variances across clusters [see e.g. \\
                    & @gelm06, @hox17 and @jong21]. \\
\hline
\end{tabular}
\end{table}

@@

The process of analyzing multilevel data is further complicated when not all data entries are observed. Just as with single level data, missingness may occur at the unit level. But with multiple levels of data comes the potential for clustered missingness. Therefore, incomplete multilevel data can be categorized into two general patterns: systematic missingness and sporadic missingness [@resc13]. Systematic missingness implies that one or more variables are never observed in a certain cluster. With sporadic missingness there may be observed data for some but not all units in a cluster [@buur18; @jola18]. We have visualized this difference in Figure 1, which shows an $n \times p$ set $\mathbf{X} = X_1, \dots, X_p$, with $n$ units distributed over $N$ clusters and $p$ variables. Column $X_1$ is completely observed, column $X_2$ is systematically missing in cluster 2, and column $X_3$ is sporadically missing. To analyze these incomplete data, we have to take the nature of the missingness and the cluster structure into account. For example, the sporadic missingness in $X_3$ could be easily amended if this would be a cluster-level variable (and thus constant within clusters). We could then just extrapolate the true (but missing) value of $X_3$ for unit 1 from unit 2, and the value for unit 4 from unit 3. If $X_3$ would instead be a unit-level variable (which may vary within clusters), we could not just recover the unobserved 'truth', but would need to use some kind of missing data method, or discard the incomplete units altogether (i.e., complete case analysis). Complete case analysis can however introduce bias in statistical inferences and lowers statistical power. Further, with the systematic missingness in $X_2$, it would be impossible to fit a multilevel model without accommodating the missingness in some way. Complete case analysis in that case would mean excluding the entire cluster from the analyses. The wrong choice of missing data handling method can thus be extremely harmful to the inferences.

<!-- Obviously, excluding observations is not a desirable workflow.  -->

```{r patterns, fig.height=2.5, fig.width=4.1, fig.cap = "Missingness in multilevel data", echo=FALSE}
# blankcol <- c(rep("", 4), "...", "")
# dat <- expand.grid(rows = 1:6, cols = 1:6) %>% 
#   cbind(text = c("1", "1", "2", "2", "...", "N", rep(blankcol, 3), rep("...", 6), blankcol),
#         miss = c(rep("", 14), "NA", "NA", "", "", "NA", "", "", "NA", rep("", 14)))
dat <- expand.grid(rows = 1:6, cols = 1:6) %>% 
  cbind(text = c("1", "1", "2", "2", "", "N", rep("", 30)),
        miss = c(rep("", 14), "NA", "NA", "", "", "NA", "", "", "NA", rep("", 14)))

ggplot(dat, aes(x = cols, y = rows)) +
  geom_tile(fill = "white", color = "black") +
  geom_text(aes(label = text), color = "black") +
  geom_text(aes(label = miss), color = mice:::mdc(2), family = "mono", fontface = "bold") + 
  scale_x_continuous(breaks = 1:6, labels = c("cluster", expression(X[1]), expression(X[2]), expression(X[3]), "...", expression(X[p])), name = NULL, position = "top") +
  scale_y_continuous(breaks = 1:6, labels = c(1:4, "...", "n"), name = NULL, trans = "reverse") +
  theme_minimal() +
  theme(panel.grid = element_blank())
```

@@

<!-- Another characteristic of the missing data to take into account in analyses is the mechanism behind the missingness. Although the essence of the true non-response mechanism may not be known, it can be inferred or assumed to be one of the following: -->
<!-- - Missing Completely At Random (MCAR), where the probability to be missing is equal across all data entries; -->
<!-- - Missing At Random (MAR), where the probability to be missing depends on observed information; -->
<!-- - Missing Not At Random (MNAR), where the probability to be missing depends on unrecorded information, making the missingness non-ignorable [@rubi76; @meng94]. -->
<!-- Depending on the assumed missingness mechanism, missing data handling strategies may be more or less suitable, see e.g., @yuce08 and @hox15. -->
<!-- "We find it useful to distinguish the missingness pattern, which describes which values are missing and observed in the data matrix and the missingness mechanism (or mechanisms), which concerns the relationship between missingness and the values of variables in the data matrix" (Little and Rubin, p. 8). -->

@@

This papers serves as a tutorial for imputing incomplete multilevel data with \pkg{mice}. We provide practical guidelines and code snippets for different missing data situations, including missing not at random (MNAR) mechanisms [where the probability to be missing depends on unrecorded information, making the missingness non-ignorable, @rubi76; @meng94]. For reasons of brevity, we focus on imputation by chained equations wit \pkg{mice} exclusively^[Note that the alternative, joint modeling imputation for multilevel data or \pkg{jomo} @jomo, has been implemented in \pkg{mice} as well but is outside the scope of this tutorial.]. Other useful resources for the analysis of incomplete multilevel data include the \proglang{R} packages \pkg{mitml}, \pkg{miceadds}, and \pkg{mdmb}, and empirical work by @audi18 and @grun18. Please note that this tutorial paper assumes a basic level of knowledge on multilevel models.^[Note to self: We're providing an overview of implementations. It's up-to the reader to decide which multilevel strategy suits their data. We won't go into detail for the different methods (and equations). This paper is just a software tutorial, so we'll keep it practical.] Assumed knowledge also includes the use of the 'piping operator', `%>%`, adopted from the \pkg{magrittr} package, and the \pkg{lme4} notation for multilevel models.^[TODO: Add environment info, seed and version number(s) somewhere!]

We illustrate how to impute incomplete multilevel data by means of three case studies:

- `popmis` from the \pkg{mice} package [simulated data on perceived popularity, $n = 2,000$ pupils across $N = 100$ schools, @mice];
- `hiv` from the \pkg{GJRM} package [simulated data on HIV diagnoses, $n = 6,416$ patients across $N = 9$ regions, @GJRM];
- `impact` from the \pkg{metamisc} package [empirical data on traumatic brain injuries, $n = 11,022$ patients across $N = 15$ studies, @metamisc].

For each of these datasets, we will discuss the nature of the missingness, choose one or more imputation models and evaluate the imputed data, but we will also highlight one specific aspect of the imputation workflow. With the `popmis` data, we show how (and how not) to develop an imputation model. With the `hiv` data we focus on extending the imputation model to include Heckman-type selection-inclusion methods. With the `impact` data we provide an example of multivariate missingness in real-world data. Together, this should give enough scaffolding for applied researchers who are faced with incomplete multilevel data.^[TODO: Add notation paragraph or 'translation table' linking multilevel equations to \pkg{lme4} formulas. Use betas instead of gamma's and mu's. Add interpretation of values in predictormatrix (-2 for the cluster variable, 2 for random effects). Add ICC and congeneality here as well. And make missingness mechanism table as well.]

@@

Set-up the R environment and load the necessary packages:

```{r}
set.seed(2022)
library(mice)
library(ggmice)
library(ggplot2)
library(dplyr)
library(lme4)
library(mitml)
```

<!-- --- -->

<!-- \begin{center} -->
<!-- REVIEW UNTIL THIS POINT SVP -->
<!-- \end{center} -->

<!-- --- -->

# Case Study I: How (not) to impute

```{r pop, echo=FALSE, message=FALSE, warning=FALSE}
# estimates complete popularity data
est_comp <- lme4::lmer(popular ~ 1 + sex + texp + sex:texp + (1 + sex | school), popcomp) 
# fixed effects only
est_comp_fixed <- est_comp %>% 
  broom.mixed::tidy("fixed", conf.int = TRUE) %>% 
  mutate(term = term,
         estimate  = paste0(
           round3(estimate), " [", 
           round3(conf.low), ", ", 
           round3(conf.high), "]"), 
         .keep = "none") %>% 
  as.data.frame()
# # random effects
# est_comp %>% 
#   broom.mixed::tidy("ran_pars") %>% 
#   mutate(term = term, #factor(term, levels = unique(.$term), labels = c()),
#          estimate  = round3(estimate), 
#          .keep = "none")
# # rename broom.mixed estimates
# lvl = c("(Intercept)", "sex", "texp", "sex:texp", "sd__(Intercept)", "cor__(Intercept).sex", "sd__sex", "sd__Observation")
# lbl = c("Intercept", "Sex", "Teacher experience", "Cross-level interaction", "SD cluster-level residual error", "Cor. cluster-level error and slopes sex", "SD residual slopes sex", "SD unit-level residual error")

# compare complete data to missingness
# popcomp <- popcomp %>% 
#   cbind(R = is.na(popmis$popular))
# t.test(x = popcomp$popular[popcomp$R], y = popcomp$popular[!popcomp$R]) 
# ggplot(popcomp, aes(y=popular, x = R)) +
#   geom_boxplot()
```

In this section we'll go over the different steps involved with imputing incomplete multilevel data. The data we're using is the `popmis` dataset from the `mice` package. This is a simulated dataset with pupils ($n = 2000$) clustered within schools ($N = 100$). In this tutorial we'll use the following variables:

  - `school`,     school identification number (clustering variable);
  - `popular`,    pupil popularity (self-rating between 0 and 10; unit-level);
  - `sex`,	      pupil sex (0=boy, 1=girl; unit-level);
  - `texp`,	      teacher experience (in years; cluster-level).

The analysis model corresponding to this dataset is multilevel regression with random intercepts, random slopes and a cross-level interaction. The outcome variable is `popular`, which is predicted from the unit-level variable `sex` and the cluster-level variable `texp`. The regression equation^[add the 'level notation' (Bryk and Raudenbush, 1992) and/or matrix notation ('linear mixed effects model'; Laird and Ware, 1982) too?] and `lme4` notation for this model are

$$
\text{popular}_{ij} =
\gamma_{00} + 
\gamma_{10} \text{ sex}_{ij} + 
\gamma_{01} \text{ texp}_{j} + 
\gamma_{11} \text{ texp}_{j} \times \text{sex}_{ij} + 
u_{0j} + 
u_{1j} \text{ sex}_{ij} + 
e_{ij} \\
$$

$$
\texttt{popular} \sim  \texttt{1 + sex + texp + sex:texp + (1 + sex | school)}
$$

<!-- Intercept-only model: lmer(formula = popular ~ 1 + (1|class)) -->
<!-- First level predictors: lmer(formula = popular ~ 1 + sex + (1|class)) -->
<!-- First and second level predictors: lmer(popular ~ 1 + sex + texp + (1 | class)) -->
<!-- First and second level predictors with random slopes: lmer(formula = popular ~ 1 + sex + texp + (1 + sex | class) -->
<!-- First and second level predictors with random slopes and cross-level interaction: lmer(formula = popular ~ 1 + sex + texp + sex:texp + (1 + sex | class)) -->

Since the data is simulated and the missingness is induced, we can compare our inferences after imputation to the true complete data. The data is created in such a way that the clustering variable `school` explains quite some variance in the outcome variable `popular`. We express this using the intraclass correlation,  ICC $=$ `r round(icc(popular ~ as.factor(school), data = popcomp), 2)`. We'll evaluate the ICC after each missing data strategy, and compare the estimated fixed effects:

```{r comp_est, echo=FALSE}
est_comp_fixed %>% setNames(c("", "Estimate with 95% CI"))
# TODO: make this a figure with the different methods as facets
```

### Incomplete data

Load the data into the environment and select the relevant variables:

```{r}
popmis <- popmis[, c("school", "popular", "sex", "texp")] 
```
Plot the missing data pattern:

```{r pop_pat, fig.cap = "Missing data pattern in the popularity data"}
plot_pattern(popmis)
```

The missingness is univariate and sporadic, which is illustrated in the missing data pattern in Figure \ref{fig:pop_pat}. The ICC in the incomplete data is `r round(icc(popular ~ as.factor(school), data = na.omit(popmis)), 2)`. This tells us that the multilevel structure of the data should probably be taken into account. If we don't, we'll may end up with incorrect imputations, biasing the effect of the clusters towards zero.

Plot the correlations in the incomplete data:

```{r pop-corr}
plot_corr(popmis)
```

To develop the best imputation model for the incomplete variable `popular`, we need to know whether the missingness depends on the observed values of other variables. We'll highlight one other variable to illustrate, but ideally one would inspect all relations. The questions we'll ask are: 'Does the missing data of pupil popularity (`popular`) depend on observed teacher popularity (`texp`)?'. This can be evaluated statistically, but visual inspection usually suffices. We'll make a histogram of `texp` separately for the pupils with known popularity and missing popularity.

Plot the histogram for teacher experience conditional on the missingness indicator of `popular`:

>>>>

---
documentclass: jss
author:
  - name: Hanne Oberman
    affiliation: Utrecht University
    address: |
      | Padualaan 14
      | 3584 CH Utrecht
    email: \email{h.i.oberman@uu.nl}
    url: https://hanneoberman.github.io/
  - name: "Johanna Munoz Avila"
    affiliation: University Medical Center Utrecht \AND
  - name: Valentijn de Jong
    affiliation: University Medical Center Utrecht 
  - name: Gerko Vink
    affiliation: Utrecht University \AND
  - name: Thomas Debray
    affiliation: University Medical Center Utrecht 
title:
  formatted: "Imputation of Incomplete Multilevel Data with \\pkg{mice}"
  plain:     "Imputation of Incomplete Multilevel Data with mice"
  short:     "Multilevel \\pkg{mice}"
abstract: >
  This is a tutorial paper on imputing incomplete multilevel data with \pkg{mice}. Footnotes in the current version show work in progress/under construction. The last section is not part of the manuscript, but purely for reminders. We aim to submit at JSS, so there is no word count limit ("There is no page limit, nor a limit on the number of figures or tables"). [Just adding some text to get a better guess of what the actura abstract will look like: Lorem ipsum dolor sit amet, consectetur adipiscing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua. Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat. Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur. Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.]
keywords:
  # at least one keyword must be supplied
  formatted: [missing data, multilevel, clustering, "\\pkg{mice}", "\\proglang{R}"]
  plain:     [missing data, multilevel, clustering, mice, R]
preamble: >
  \usepackage{amsmath}
header-includes: 
 - \usepackage{graphicx} 
 - \usepackage{mathtools}
 - \usepackage{ulem}
output: 
    rticles::jss_article:
      keep_tex: yes
      number_sections: yes
    # word_document: default
bibliography: ../References/multilevelmice.bib
editor_options: 
  chunk_output_type: inline
---

@@

# environment
set.seed(123)
library(dplyr)
library(ggplot2)
```

<!-- Missing data pattern with clustering -->
<!-- TODO: simplify argument in md pattern plot and/or network plot for md patterns (summarized per variable?) -->
<!-- TODO: test if there's a random distribution of missingness in the md pattern? -->
<!-- TODO: check spss missing data pie chart? -->
<!-- TODO: similarity of patterns over clusters -->
<!-- TODO: gradient (alpha) to show in how many of the clusters the row in the pattern occurs -->
<!-- TODO: 3 columns could be 3 boxplots, over these clusters we have a distribution of these statistics across clusters.  -->
<!-- TODO: look at vim -->
<!-- TODO: check if shepley plot is an option -->

<!-- Conditional plots missingness indicator -->
<!-- TODO: check if up-side-down plot works -->
<!-- TODO: check smoothing in geom_density function and make it the inverse of the sample size -->
<!-- TODO: add functions to mice -->
<!-- TODO: add title and informative legend (pop obs and pop miss) -->
<!-- TODO: add facets for some clusters, or add propensity score distribution `is.na(popular) ~ .` -->
<!-- TODO: make it average of cluster dens, not marginal, alternatively add a quartile line around the density with geom_ribbon -->
<!-- TODO: add fill and/or some way of expressing how much the densities overlap (overlap coefficient in %?) -->

@@

In many contemporary data analysis efforts, some form of hierarchical or clustered data structures are recorded. In the simplest case, such a structure entails the nesting of units within clusters (e.g., students within school classes). More complex clustered structures may occur when there are multiple hierarchical levels (e.g., patients within hospitals within regions or countries), or when the clustering is non-nested (e.g., electronic health record data from diverse settings and populations within large databases). The clustered structure of multilevel data should be taken into account when developing analysis models: 1) for the simple reason that groups of observations share some common variance, and 2) because ignoring multilevel structures can be harmful to the statistical inferences and introduce bias in estimators [@hox17]. There are many names for models that take clustering into account. Some popular examples are 'multilevel models', 'hierarchical models', 'mixed effect models' and 'random effect models'. Table \ref{tab:clus} provides an overview of some key concepts in multilevel modeling.

\begin{table}[tb]
\caption{Concepts in multilevel methods}
\label{tab:clus}
\centering
\begin{tabular}{ll}
\hline
\textbf{Concept} & \textbf{Details}   \\
\hline
ICC                 & The variability due to clustering is often measured by means of the \\
                    & intraclass coefficient (ICC). The ICC can be seen as the percentage \\
                    & of variance that can be attributed to the cluster-level, where a high \\
                    & ICC would indicate that a lot of variability is due to the cluster \\
                    & structure. \\
Random intercept    & Multilevel models typically accommodate for variability by including \\
                    & a separate group mean for each cluster. In addition to random \\
                    & intercepts, multilevel models can also include random effects and \\
                    & heterogeneous residual error variances across clusters [see e.g. \\
                    & @gelm06, @hox17 and @jong21]. \\
\hline
\end{tabular}
\end{table}

@@

The process of analyzing multilevel data is further complicated when not all data entries are observed. Just as with single level data, missingness may occur at the unit level. But with multiple levels of data comes the potential for clustered missingness. Therefore, incomplete multilevel data can be categorized into two general patterns: systematic missingness and sporadic missingness [@resc13]. Systematic missingness implies that one or more variables are never observed in a certain cluster. With sporadic missingness there may be observed data for some but not all units in a cluster [@buur18; @jola18]. We have visualized this difference in Figure 1, which shows an $n \times p$ set $\mathbf{X} = X_1, \dots, X_p$, with $n$ units distributed over $N$ clusters and $p$ variables.

```{r patterns, fig.height=1.5, fig.width=4.1, fig.cap = "Missingness in multilevel data", echo=FALSE}
dat <- expand.grid(rows = 1:6, cols = 1:6) %>% 
  cbind(text = c("1", "1", "2", "2", "", "N", rep("", 30)),
        miss = c(rep("", 14), "NA", "NA", "", "", "NA", "", "", "NA", rep("", 14)))
ggplot(dat, aes(x = cols, y = rows)) +
  geom_tile(fill = "white", color = "black", size = 0.5) +
  geom_text(aes(label = text), color = "black", size = 3) +
  geom_text(aes(label = miss), color = mice:::mdc(2), family = "mono", fontface = "bold") + 
  scale_x_continuous(
    breaks = 1:6,
    labels = c("cluster", expression(X[1]), expression(X[2]), expression(X[3]), "...", expression(X[p])),
    name = NULL,
    position = "top"
  ) +
  scale_y_continuous(breaks = 1:6, labels = c(1:4, "...", "n"), name = NULL, trans = "reverse") +
  coord_cartesian(expand = c(0,0,0,0)) +
  theme_minimal() +
  theme(panel.grid = element_blank(),
        plot.margin = unit(c(0,0,0,0), "pt"))
```
Column $X_1$ in Figure 1 is completely observed, column $X_2$ is systematically missing in cluster 2, and column $X_3$ is sporadically missing. To analyze these incomplete data, we have to take the nature of the missingness and the cluster structure into account. For example, the sporadic missingness in $X_3$ could be easily amended if this would be a cluster-level variable (and thus constant within clusters). We could then just extrapolate the true (but missing) value of $X_3$ for unit 1 from unit 2, and the value for unit 4 from unit 3. If $X_3$ would instead be a unit-level variable (which may vary within clusters), we could not just recover the unobserved 'truth', but would need to use some kind of missing data method, or discard the incomplete units altogether (i.e., complete case analysis). Complete case analysis can however introduce bias in statistical inferences and lowers statistical power. Further, with the systematic missingness in $X_2$, it would be impossible to fit a multilevel model without accommodating the missingness in some way. Complete case analysis in that case would mean excluding the entire cluster from the analyses. The wrong choice of missing data handling method can thus be extremely harmful to the inferences.

A key characteristic of the missing data to take into account in analyses is the mechanism behind the missingness. We distinguish between MCAR, MAR and MNAR in theory (see Table \ref{tab:miss}), but in practice this distinction is less clear. Since the essence of the true non-response mechanism may not be known, it is generally inferred or assumed to be ignorable (i.e., MCAR or MAR). [TODO: add that this assumption may not always be valid, especially with modern types of big dasta sources.] Depending on the actual missingness-generating mechanism, missing data handling strategies may be more or less suitable, see e.g., @yuce08 and @hox15.

\begin{table}[tb]
\caption{Concepts in missing data methods}
\label{tab:miss}
\centering
\begin{tabular}{ll}
\hline
\textbf{Concept} & \textbf{Details}   \\
\hline
MCAR    & Missing Completely At Random, where the probability to be missing is equal \\
& across all data entries \\
MAR     & Missing At Random, where the probability to be missing depends on observed \\
& information \\
MNAR    & Missing Not At Random (MNAR), where the probability to be missing \\
& depends on unrecorded information, making the missingness non-ignorable \\
& [@rubi76; @meng94]. \\
& [TODO: add congeniality.] \\
\hline
\end{tabular}
\end{table}

@@

[TODO: clarify why clustering is relevant during imputation, and why this exposes the need for specialized imputation methods and more attention during their implementation ("thou shall not simply run `mice()` on any incomplete dataset"). And add overview of possible predictor matrix values.]

@@

This papers serves as a tutorial for imputing incomplete multilevel data with \pkg{mice} in \proglang{R}. We provide practical guidelines and code snippets for different missing data situations, including non-ignorable mechanisms. For reasons of brevity, we focus on multilevel imputation by chained equations with \pkg{mice} exclusively; other imputation methods and packages (e.g., \pkg{jomo} and \pkg{mdmb}) are outside the scope of this tutorial. Assumed knowledge includes basic familiarity with multilevel imputation [see e.g. @audi18 and @grun18] and the \pkg{lme4} notation for multilevel models (see Table \ref{tab:mod}).

\begin{table}[tb]
\caption{Concepts in missing data methods}
\label{tab:mod}
\centering
\begin{tabular}{ll}
\hline
\textbf{Concept} & \textbf{Details}   \\
\hline
& [TODO: explain \pkg{lme4} notation here] \\
\hline
\end{tabular}
\end{table}

# Case study

We illustrate how to impute incomplete multilevel data by means of a case study: `impact` from the \pkg{metamisc} package [empirical data on traumatic brain injuries, $n = 11,022$ patients across $N = 15$ studies, @metamisc]. `impact` is traumatic brain injury data with patients, $n = 11022$, clustered in studies, $N = 15$. With the following 11 variables:

[TODO: add more info about the complete data.] The analysis model for this dataset is a prediction model with `mort` as the outcome. In this tutorial we'll estimate the adjusted prognostic effect of `ct` on unfortunate outcomes. The estimand is the adjusted odds ratio for `ct`, after including `type`, `age` `motor_score` and `pupil` into the analysis model: 
```{r mod}
mod <- mort ~ 1 + type + age + motor_score + pupil + ct + (1 | name)
```

Note that variables `hypots`, `hypox`, `tsah` and `edh` are not part of the analysis model, and may thus serve as auxiliary variables for imputation.

## Setup

[TODO: Add environment info, seed and version number(s) somewhere.] Set up the R environment and load the necessary packages:
```{r env, message=FALSE, warning=FALSE}
set.seed(2022)
library(mice)         # for imputation
library(ggmice)       # for visualization
library(ggplot2)      # for visualization
library(dplyr)        # for data wrangling
library(lme4)         # for multilevel modeling
library(broom.mixed)  # for multilevel estimates
library(mitml)        # for multilevel pooling
```

The `impact` data included in the \pkg{metamisc} package is a complete data set. The original data has already been imputed once (Steyerberg et al, 2008). For the purpose of this tutorial we have induced missingness (mimicking the missing data in the original data set before imputation). The resulting incomplete data can be accessed from [zenodo link to be created](https://zenodo.com).

Load the complete and incomplete data into the R workspace:
```{r data, eval = FALSE}
data("impact", package = "metamisc")
dat <- read.table("link/to/the/data.txt")
```

```{r data-actually, include=FALSE}
data("impact", package = "metamisc")
dat <- readRDS("../Data/impact_incomplete.RDS")
```

We will use the following estimates as comparative truth in this tutorial [TODO: make this a table or forest plot instead, to see if there is heterogeneity in the association of `ct` with `mort`]:
```{r truth}
fit <- glmer(mod, family = "binomial", data = impact) 
tidy(fit, conf.int = TRUE, exponentiate = TRUE) 
```

[TODO: add ICC before/after imputation and interpret: This tells us that the multilevel structure of the data should probably be taken into account. If we don't, we'll may end up with incorrect imputations, biasing the effect of the clusters towards zero.]

[TODO: add descriptive statistics of the complete and incomplete data.]

## Missingness

To explore the missingness, it is wise to look at the missing data pattern:
```{r pattern, fig.height=7.1}
plot_pattern(dat, rotate = TRUE)
```

This shows... [TODO: fill in that we need to impute `ct` and `pupil`.]

To develop the best imputation model, we need to investigate the relations between the observed values of the incomplete variables and the observed values of other variables, and the relation between the missingness indicators of the incomplete variables and the observed values of the other variables. To see whether the missingness depends on the observed values of other variables, we... [TODO: fill in that we can test this statistically or use visual inspection (e.g. a histogram faceted by the missingness indicator).]

We should impute the variables `ct` and `pupil` and any auxiliary variables we might want to use to impute these incomplete analysis model variables. We can evaluate which variables may be useful auxiliaries by plotting the pairwise complete correlations:
```{r impact_corr}
plot_corr(dat, rotate = TRUE)
```

This shows us that `hypox` and `hypot` would not be useful auxiliary variables for imputing `ct`. Depending on the minimum required correlation, `tsah` could be useful, while `edh` has the strongest correlation with `ct` out of all the variables in the data and should definitely be included in the imputation model. For the imputation of `pupil`, none of the potential auxiliary variables has a very strong relation, but `hypots` could be used. We conclude that we can exclude `hypox` from the data, since this is neither an analysis model variable nor an auxiliary variable for imputation:
```{r}
dat <- select(dat, !hypox)
```

<!-- We'll highlight one other variable to illustrate, but ideally one would inspect all relations. The questions we'll ask are: 'Does the missing data of pupil popularity (`popular`) depend on observed teacher popularity (`texp`)?'. This can be evaluated statistically, but visual inspection usually suffices. We'll make a histogram of `texp` separately for the pupils with known popularity and missing popularity. -->

## Complete case analysis

As previously stated, complete case analysis lowers statistical power and may bias results. The complete case analysis estimates are:
```{r cca}
fit <- glmer(mod, family = "binomial", data = na.omit(dat)) 
tidy(fit, conf.int = TRUE, exponentiate = TRUE) 
```

As we can see... [TODO: fill in.]

## Imputation model

@@

The first imputation model that we'll use is likely to be invalid. We do *not* use the cluster identifier `name` as imputation model predictor. With this model, we ignore the multilevel structure of the data, despite the high ICC. This assumes exchangeability between units. We include it purely to illustrate the effects of ignoring the clustering in our imputation effort. We'll use the default imputation methods in `mice()` (predictive mean matching to impute the continuous variables and logistic regression to impute binary variables).

---

\begin{center}

Updated until here!

\end{center}

---

Create a methods vector and predictor matrix, and make sure `name` is not included as predictor:
```{r ignore}
meth <- make.method(dat) # methods vector
pred <- quickpred(dat)   # predictor matrix
plot_pred(pred)
```

<!-- We'll now use `school` as a predictor to impute all other variables. This is still not recommended practice, since it only works under certain circumstances and results may be biased [@drec15; @ende16]. But at least, it includes some multilevel aspect. This method is also called 'fixed cluster imputation', and uses N-1 indicator variables representing allocation of N clusters as a fixed factor in the model [@reit06; @ende16]. Colloquially, this is 'multilevel imputation for dummies'.  -->

<!-- Add: doesn't work with syst missing (only sporadically). There's some pro's and con's. May not differ much if the number of clusters is low. -->

<!-- The more the random effects are of interest, the more you need ml models. -->

```{r impact-pred, include=FALSE, eval=FALSE}
pred <- quickpred(dat)
plot_pred(pred)
